Further Evidence of Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2022-02-02 src/ms.tex
JVS comments

@@

\subsection{Where do the pile-ups end?} \label{subsec:extent}
It appears from Figures~\ref{fig:kde} and \ref{fig:models} that the number density of stars on the long- and short-period pile-ups declines towards cooler \teff. While it not clear whether or not these declines are astrophysical in nature, the result of the Kepler selection function or observational biases, or some combination of effects, we attempted to characterize the extent of the pile-ups through the following approach. We found through inspection that constant Rossby curves of Ro=0.5 and Ro=1.3 (for the short- and long-period pile-ups, respectively) appear to describe the highest density contours found from Gaussian kernel density estimation of the \lamostmcq sample. In windows of 10~K width we measured the fraction of stars with periods within 1~d of each of the two constant Rossby curves, relative to the total number of stars in that \teff window. We found that the relative fraction of stars on the long-period pile-up declines rapidly between 6500~K and 5800~K, by more than half in that temperature range before plateauing at cooler \teff (Figure~\ref{fig:fraction}). The relative fraction of stars on the short-period edge declines more slowly, and below the temperature of the Sun, the relative fractions of stars on the two pile-ups are nearly equal. Thus, it is not yet clear if either pile-up extends to temperatures cooler than \tsun.

@@

As discussed in \S\ref{subsec:models} and \S\ref{subsec:rossby}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}. {\color{red} Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The present samples have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is less biased and more likely to contain pile-up stars that our sample may have missed. It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offsets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered clearly indicates that the long-period pile-up is apparent for stars with \teff~$\gtrsim$~\teffmin with Rossby numbers lower than those associated with a detection edge.}

>>>>

%Stars reach Ro = 2 at very roughly halfway through the MS, and stay there until the MS turnoff. For higher mass stars with MS lifetimes of 4 Gyr, it means that stars on the ridge should have a ~2 Gyr age spread. For stars with 10 Gyr MS lifetimes, it should be something close to a 5 Gyr age spread.

@@

\subsection{Where do the pile-ups end?} \label{subsec:extent}
It appears from Figures~\ref{fig:kde} and \ref{fig:models} that the number density of stars on the long- and short-period pile-ups declines towards cooler \teff, as predicted by the WMB model \citep[see Figure 13 of][]{vanSaders2019}. While it not clear whether or not these declines are astrophysical in nature, the result of the Kepler selection function or observational biases, or some combination of effects, we attempted to characterize the extent of the pile-ups through the following approach. We found through inspection that constant Rossby curves of Ro=0.5 and Ro=1.3 (for the short- and long-period pile-ups, respectively) appear to describe the highest density contours found from Gaussian kernel density estimation of the \lamostmcq sample. In windows of 10~K width we measured the fraction of stars with periods within 1~d of each of the two constant Rossby curves, relative to the total number of stars in that \teff window. We found that the relative fraction of stars on the long-period pile-up declines rapidly between 6500~K and 5800~K, by more than half in that temperature range before plateauing at cooler \teff (Figure~\ref{fig:fraction}). The relative fraction of stars on the short-period edge declines more slowly, and below the temperature of the Sun, the relative fractions of stars on the two pile-ups are nearly equal. Thus, it is not yet clear if either pile-up extends to temperatures cooler than \tsun.

@@

As discussed in \S\ref{subsec:models} and \S\ref{subsec:rossby}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}. {\color{red} Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The present samples have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is not biased against quiet, unspotted stars and is more likely to contain pile-up stars that our sample may have missed. It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offsets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered clearly indicates that the long-period pile-up is apparent for stars with \teff~$\gtrsim$~\teffmin with Rossby numbers lower than those associated with a detection edge.}
